The Inconsistency in G\"odel's Ontological Argument: A Success Story for Artificial Intelligence in Metaphysics (Extended Abstract)

### 2016-07-08 Papers/2016/KI/main.tex
update

\title{The Inconsistency in G\"odel's Ontological Argument: A Success
  Story for Automated Theorem Proving in Metaphysics}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  We discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in the early 1970's, the
  inconsistency of the axioms used in the argument remained unnoticed
  until 2013, when it was detected automatically by the higher-order
  theorem prover \textsc{Leo-II}.  Understanding and verifying the
  refutation generated by the prover turned out to be a time-consuming
  task. Its completion, as reported here, required the
  reconstruction of the refutation in the Isabelle proof assistant.% , and it
  % also led to a novel and more efficient way of automating
  % higher-order modal logic \textbf{S5} with a universal accessibility
  % relation. Furthermore, the development of an improved syntactical
  % hiding for the utilized logic embedding technique allows the refutation to be
  % presented in a human-friendly way, suitable for non-experts in the
  % technicalities of higher-order theorem proving. This brings us a
  % step closer to wider adoption of logic-based artificial intelligence
  % tools by philosophers.

  (This paper is a short version of our contribution to IJCAI 2016.)
\end{abstract}

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticize, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\cite{sobel2004logic} and Oppy~\cite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when we~\cite{J30,C40} first
reported our successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
our work has received a media repercussion on a global scale\footnote{A
  collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf.~\cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf.~\cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in our preceding work.

The main novel contribution is a detailed
analysis (in various modal logics) of the inconsistency of G\"{o}del's
original version of the axioms used in his manuscript
\cite{GoedelNotes}. The extraction, reconstruction and
verification of an informal, human intuitive argument has been an open
problem since the first detection of this inconsistency by
the
\textsc{Leo-II} prover \cite{C40}.  The verified refutation (discussed in
\S\ref{sec:inconsistency}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically profound
and never presented in the literature. The detection of this
inconsistency in combination with the work reported here thus
demonstrates that 
higher-order automated theorem provers are capable of assisting in
the discovery and elucidation of \emph{new} and philosophically
relevant knowledge.

@@

\paragraph{Related Work.}
First successful applications of theorem proving technology in
metaphysics were reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on,
Rushby~\cite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\cite{Proslogion} simpler 
and older ontological argument. In contrast, the greater complexity of G\"odel's argument requires the formalization and automation of variants of \emph{higher-order} and \emph{modal} logics.

\begin{wrapfigure}[22]{r}{0.55\textwidth} \centering \vskip-3em
  \includegraphics[width=.52\textwidth]{./Scott_S5U.png}
  \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{wrapfigure} 
\section{A Brief History}\label{sec:history}
St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz~\cite{Adams}.  
Although G\"odel's notion of positive property is not exactly the same
as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). G\"odel's
different definition of essence can be seen 
% either in his manuscript
% (Fig.~\ref{GoedelScript}) or, in more modern notation,
in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

\begin{wrapfigure}[20]{r}{0.55\textwidth}
    \centering \vskip-1em
    \centerline{\includegraphics[width=0.52\textwidth]{./InconsistencyIsabelleK.png}}
 \caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{wrapfigure}
G\"odel's omission has been
considered inessential and merely an oversight by many. 
% For instance,
% Hazen [\cite[p.365]{Hazen}] states that ``G\"odel left this
% clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
\cite[p.364]{oppy00:_respon_gettin};
\cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  at the time that G\"odel's axioms were inconsistent.}

@@

For a computational analysis of G\"odel's ontological argument, we
utilised an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. This idea has already been successfully
employed in earlier work \cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers. The embedding
approach is, therefore, very flexible.

@@

@@

The axioms and definitions in G\"odel's manuscript are inconsistent;
this was detected automatically by the prover
\textsc{Leo-II}. Here we presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic \K.

We have also presented several technical improvements regarding the
semantic embedding approach. In particular, we have achieved a
nearly perfect match between pen and paper presentations in HOML and
the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
is now fully transparent, more user-friendly and ready for wider adoption.

On the other hand, there is still room for many pragmatical
improvements in Isabelle; just one example: in default setting,
Sledgehammer does not immediately inform the user when a proof has
been found and instead silently first executes a series of
time-consuming proof analysis processes (e.g. its dependency
minimization), before it eventually reports success. For G\"odel's
theorem T3 (\textit{Necessarily, there exists God}), for example, this
phase of silence takes several minutes --- during which the user might
actually give up on the proof attempt --- even though \textsc{Leo-II} already
reported success to Sledgehammer after 2.5 seconds.

More importantly, our work reveals a challenge for automated reasoning:
the (so far partially manual) extraction of an informal argument from a formal proof. 
Without accompanying human-understandable explanations,
the proofs generated by provers such as \textsc{Leo-II} or Metis, will
presumably be only of limited value for philosophers, for whom intuitive
arguments remain crucial for the acceptance of novel results.

Another open problem that we solved in this paper is a fully automatic
proof of T3 directly from Scott's axioms. Again, this proof was
contributed by \textsc{Leo-II}. This has become possible only
after we provided a more efficient embedding for HOML \SFiveU (instead of \SFive) in HOL.

Both the automated detection of the inconsistency in G\"odel's axioms
and the fully automatic proof of T3 from Scott's axioms demonstrate
the potential of our AI technology for philosophy: this technology is,
in its current state of development, already capable of contributing novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.  
\vfill

\noindent\textbf{Acknowledgments:} We thank Chad Brown, who contributed to the rational  reconstruction
of the inconsistency argument.

@@

\small
\bibliographystyle{plain}
\bibliography{Bibliography}

>>>>

\title{The Inconsistency in G\"odel's Ontological Argument: A Success
  Story for Artificial Intelligence in Metaphysics (Extended Abstract)}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  We discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of G\"odel's argument this
  inconsistency remained unnoticed until 2013, when it was detected
  automatically by the higher-order theorem prover
  \textsc{Leo-II}.  % Understanding and verifying the
  % refutation generated by the prover turned out to be a time-consuming
  % task. Its completion, as reported here, required the
  % reconstruction of the refutation in the Isabelle proof assistant.% , and it
  % also led to a novel and more efficient way of automating
  % higher-order modal logic \textbf{S5} with a universal accessibility
  % relation. Furthermore, the development of an improved syntactical
  % hiding for the utilized logic embedding technique allows the refutation to be
  % presented in a human-friendly way, suitable for non-experts in the
  % technicalities of higher-order theorem proving. This brings us a
  % step closer to wider adoption of logic-based artificial intelligence
  % tools by philosophers.

  (This text is a significantly shortened version of our contribution to IJCAI 2016.)
\end{abstract}

\section{Introduction}\label{sec:introduction}
% Without exaggeration 
Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticize, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\cite{sobel2004logic} and Oppy~\cite{sep-ontological-arguments} and the
references therein).

 % In philosophy lectures at universities the
% argument is regularly presented as a masterpiece argument in
% metaphysics. Since 2013, when we~\cite{J30,C40} first
% reported our successful initial computer-assisted
% analysis of G\"odel's proof and Scott's variant,
% our work has received a media repercussion on a global scale\footnote{A
%   collection of news articles is available at {\scriptsize
%     \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
% and numerous bloggers commented on the proof
% \cite{fuhrmann15:_blogg_goedel}.

However, G\"odel's original version of the axioms used in his
manuscript \cite{GoedelNotes} are inconsistent, and this fact has
remained unnoticed to philosophers until 2013, when it was detected
automatically by the higher-order theorem prover \textsc{Leo-II}
\cite{C40}.  We here report on the extraction, reconstruction and
verification of an informal, human intuitive argument for this
inconsistency in the Isabelle/HOL proof assistant.  The respective
presentation
of this intuitive argument further extends and improves our contribution 
to IJCAI 2016 \cite{C55}.

% The verified refutation (discussed
% in \S\ref{sec:inconsistency}) displays a surprisingly accessible
% explanation of the inconsistency, which is philosophically profound
% and never presented in the literature. 
% The detection of this
% inconsistency in combination with the work reported here thus
% demonstrates that higher-order automated theorem provers are capable
% of assisting in the discovery and elucidation of \emph{new} and
% philosophically relevant knowledge.

@@

In our experiments we have employed theorem provers for classical
higher-order logic (HOL; cf.~\cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf.~\cite{homl} and the
references therein). To bridge between the two logics we adopt the
logic embedding approach \cite{J23,C40}.

%\paragraph{Related Work.}
Applications of (first-order) theorem proving technology in
metaphysics were first reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}. Zalta also coined the term
\textit{Computational Metaphysics}. Later on, Rushby~\cite{rushby13} used the proof
assistant \textsc{PVS}. Common to both works is a
significant amount of proof-hand-coding work as well as their focus on
a non-modal formalization of St. Anselm's simpler
and older ontological argument. % In contrast, the greater complexity of
% G\"odel's argument requires the formalization and automation of
% variants of \emph{higher-order} and \emph{modal} logics.

\begin{figure}\centering 
  \includegraphics[width=.495\textwidth]{./Scott_S5U.png}
  \includegraphics[width=.495\textwidth]{./InconsistencyIsabelleK.png}
  % \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  % Axioms;  Automatic Verification of Modal Collapse} 
\caption{Scott's consistent axioms (left) and proof of the
  inconsistency of (a subset of) G\"odel's original  axioms (right)}
  % Axioms;  Automatic Verification of Modal Collapse} 
\label{Scott_Goedel}
\end{figure} 
\paragraph{A Brief History.}\label{sec:history}
% St. Anselm's ontological argument \cite{Proslogion} can be regarded as
% the ancestor of modern ontological arguments such as G\"odel's. In the
% millenium between Anselm and G\"odel, many philosophers modified and
% arguably improved Anselm's argument. Of particular importance to
% G\"odel was the work of Leibniz~\cite{Adams}.  
% Although G\"odel's notion of positive property is not exactly the same
% as Leibniz's notion of perfection, 
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}); cf. Fig.~\ref{Scott_Goedel}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

% \begin{wrapfigure}[20]{r}{0.55\textwidth}
%     \centering \vskip-1em
%     \centerline{\includegraphics[width=0.52\textwidth]{./InconsistencyIsabelleK.png}}
%  \caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
% \end{wrapfigure}
G\"odel's omission has been
considered inessential and merely an oversight by many. 
% For instance,
% Hazen [\cite[p.365]{Hazen}] states that ``G\"odel left this
% clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument.%  Especially since the discovery by Sobel
% \cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
%   $\phi\rightarrow \Box \phi$, states that contingent truth implies
%   necessary truth; it can be interpreted as \textit{everything is
%     pre-determined} or even \textit{there is no free will}. } is
% entailed by G\"odel's (or also Scott's) axioms, several variants have
% been proposed
% \cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
% attempting to avoid the modal collapse. Many of these variants omit
% the crucial conjunct in the definition of essence as well.\footnote{As
%   these variants also change other axioms, on which the inconsistency
%   of G\"odel's axioms depends, it is not necessarily the case that
%   these variants are also inconsistent; they must be analyzed
%   separately.} Opponents of the argument
% (e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
% \cite[p.364]{oppy00:_respon_gettin};
% \cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
% parodies and other criticisms, referring to variants where the
% conjunct is omitted.

@@

However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_Goedel}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  at the time that G\"odel's axioms were inconsistent.}

@@

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers. The embedding
approach is, therefore, very flexible.

@@

\textbf{Bruno: Here we should present the syntactical argument for the
  inconsistency of $\pos \nec \bot $.}

@@

The axioms and definitions in G\"odel's manuscript are inconsistent;
this was detected automatically by the prover
\textsc{Leo-II}. We presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic \K.

Extending the presentation in our IJCAI 2016 paper \cite{C55}, we have added
a syntactical proof for the inconsistency of  $\pos \nec \bot$.  Previously we had reduced the inconsistency
analysis to exactly this proposition and we had then argued
semantically that this proposition must be false already in modal
logic \K.

% We have also presented several technical improvements regarding the
% semantic embedding approach. In particular, we have achieved a
% nearly perfect match between pen and paper presentations in HOML and
% the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
% is now fully transparent, more user-friendly and ready for wider adoption.

% On the other hand, there is still room for many pragmatical
% improvements in Isabelle; just one example: in default setting,
% Sledgehammer does not immediately inform the user when a proof has
% been found and instead silently first executes a series of
% time-consuming proof analysis processes (e.g. its dependency
% minimization), before it eventually reports success. For G\"odel's
% theorem T3 (\textit{Necessarily, there exists God}), for example, this
% phase of silence takes several minutes --- during which the user might
% actually give up on the proof attempt --- even though \textsc{Leo-II} already
% reported success to Sledgehammer after 2.5 seconds.

Our work reveals a challenge for automated reasoning:
the (so far partially manual) extraction of an informal argument from a formal proof. 
Without accompanying human-understandable explanations,
the proofs generated by provers such as \textsc{Leo-II} or Metis, will
presumably be only of limited value for philosophers, for whom intuitive
arguments remain crucial for the acceptance of novel results.

% Another open problem that we solved in this paper is a fully automatic
% proof of T3 directly from Scott's axioms. Again, this proof was
% contributed by \textsc{Leo-II}. This has become possible only
% after we provided a more efficient embedding for HOML \SFiveU (instead of \SFive) in HOL.

% Both the automated detection of the inconsistency in G\"odel's axioms
% and the fully automatic proof of T3 from Scott's axioms
Our work 
demonstrates
the potential of higher-order automated theorem proving and the
semantic embeddings approach for philosophy: this technology is,
in its current state of development, already capable of contributing novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.  
\vfill

% \noindent\textbf{Acknowledgments:} We thank Chad Brown, who contributed to the rational  reconstruction
% of the inconsistency argument.

@@

\bibliographystyle{plain}
\bibliography{Bibliography}

\pagebreak
